\cursedforest\ - A Random Forest Implementation for ``Big'' and ``Wide'' Data \\ Supplementary Information

### 2016-12-19 doc/publications/plos16/plos16.tex
changes to plos16.tex

\newcommand{\cursedforest}{{\sc CursedForest}}
\newcommand{\mtry}{{\texttt mtry}}
\newcommand{\ntree}{{\texttt ntree}}

@@

The most popular random forest software packages are R-based, although current implementations do not scale well with
increasing feature size~\cite{Wright.and.Ziegle.2016}.  This is despite R now supporting long vectors (length greater
than $2^{31}-1$), and the {\sc randomForest} implementation in R lending itself to parallelization to grow multiple
trees simultaneously~\cite{Liaw.and.Weiner.2002} and them combine them.  More suitable implementations for ``wide''
genomic data have been developed in other programming languages. \textsc{CloudForest} \cite{Bressler2015} written in Go,
achieves fast run times by effective use of the CPU cache, optimizing for different classes of features and efficiently
multi-threading.  Similarly, \textsc{Ranger}~\cite{Wright.and.Ziegle.2016} written in C++ with an R front-end is
specifically optimized for dealing with large data sets.

@@

Random forests apply the technique of bootstrap aggregation or bagging to decision trees.  The training data sets are
independently drawn, with replacement, from the joint distribution of $(X,Y)$ so that each set comprises $n$
$(p+1)$-tuples $(x_1,y_1),\ldots, (x_n,y_n)$. This is done $B$ times and a tree is grown on each of these $B$ samples.
The results are then combined in an appropriate way. In a classification problem we take the majority vote for the trees
over the $q=1,..,Q$ classes,
\begin{equation*}
{{h_f}}= \arg \max_q \left(\sum_b I(h(x;B_b)=q)\right).
\end{equation*}

@@

In the case of a regression tree, the average prediction error for an individual tree $h(X; B)$ is
\begin{equation}
PE_t = E_B E_{X,Y} (Y-h(X; B))^2.
\end{equation}
Assume that, for all $B$, the tree is unbiased, i.e., $EY= E_X h(X; B)$. Then
\begin{equation}
PE_f \leq \rho PE_t
\end{equation}
where $\rho$ is the weighted correlation between residuals $(Y-h(X;B))$ and $(Y-h(X;B^\prime))$ for independent $B,B^\prime$.  
So what is required for accurate random forest regression is (i) low correlation between residuals of differing tree in
the forest, and (ii) low prediction error for the individual trees \cite{Segal.2004}. In addition to growing the trees
on bootstrapped samples, each node in the tree is split on a randomly selected subset of the variables.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% This section talks about the datasets
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Datasets}
\subsubsection{1000 Genomes Project}
We obtained the 1000 Genomes Project data as VCF files from their FTP site.  Each VCF file contains the variants for
every individual across one chromosome.  We use the phase 3 dataset which contains 2,504 individuals with approximately
81 million variant positions (i.e. a matrix of $2,504 \times 81$ million).  For compatibility with the majority of the
algorithms we demonstrate, we convert the original variant representations to double representations.  For example, we
store the absence of a variant (0\textbar0) as 0, a heterozygous variant (1\textbar0 or 0\textbar1)as 1, and a
homozygous variant (1\textbar1) as 2.  This is not the only possible encoding.  \cite{Goldstein.et.al.2011} note that of
the encoding listed in table \ref{table:encodings}, the additive encoding will present the most difficulty for a random
forest model.

@@

\subsection{Parameter settings}
We consider the parameter setting for the random forest algorithm. We use the R notation from the {\sc randomForest}
package \cite{Liaw.and.Weiner.2002} which incorporates the original Fortran code by Brieman and Cutler.  We incorporate
the advice of \cite{Liaw.and.Weiner.2002}, which we have found mirrors our own experience:
\begin{itemize}
\item \ntree\ the number of trees.  The number of trees necessary for good performance grows with the number of
  predictors.  \cite{Liaw.and.Weiner.2002} suggest that a high \ntree\ is necessary to get stable estimates of variable
  importance and proximity; however, even though the variable importance measures may vary from run to run, we note that
  it is possible for a random forest model to have a very poor fit and still have an accurate ranking of variable
  importance;
\item \mtry\ -- the number of variables considered at each split (if \mtry=$p$, we have a boosted decision
  tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger \mtry\ may give
  better performance;
\item the size and complexity of the individual trees is controlled in
  {\sc randomForest} by setting \texttt{nodesize}, the minimum size of
  terminal nodes. It is controlled in Spark ML by setting
  \texttt{maxDepth}, the maximum depth of each tree in the
  forest\footnote{The Spark ML documentation \cite[]{Spark.2016} sets
    \texttt{maxDepth} to 4 in their classification example. This may
    give an estimate with a higher bias and, as such, may be a poor
    choice. See the discussion in \cite{Dietterich.2002} where it is
    shown that bagging may reduce bias as well as variance but gives a
    better result for low bias learners. \cite{Dietterich.2002} notes
    that if the boostrap replicate approximation were correct, then
    bagging would reduce variance without changing bias.} In the TCGA
  example the RandomForests code produced trees with depths of 30+.
\end{itemize}

@@

@@

Firstly we look at CursedForest horizontal scalability for a medium size dataset of 2.5 million variables and 5000
samples, by varying \mtry\ fraction and the number of CPU cores allocated to the execution. The raw results are
presented in table \ref{synthetictimingtable} and visualized in figure \ref{figure:synth} below. Regardless of the
number of cores used, \cursedforest\ displays approximately linear dependency between the execution time and the \mtry\
parameter (presented here as a fraction of the number of variables) [Fig 1 a).

@@

Notes:
\begin{itemize}
\item we try a small simulation, a regression problem with $\beta(1:k) \sim U(1,100)$ and $y= X\beta + \epsilon$, $\epsilon \sim  N(0,\sqrt{16})$
\item \cite{Donoho.and.Stodden.2006} evaluate variable selection by 
$$\frac{\Vert\hat{\beta}-\beta\Vert_2}{\Vert\beta\Vert_2}$$
See figure \ref{figure:stepwise_lars_cv.png} 
\item 
However there is no $\hat{\beta}$ in a tree model. We are using variable importance as
a measure of the ordering of $\hat{\beta}$ (i.e. we expect variable importance to be proportional to $\hat{\beta}$). Is this correct?
\item as RF is sort of greedy approximation to combinatorial approach, it may have some success in the dead zone
\item  \cite{Donoho.and.Stodden.2006} are considering problems with $L_2$ errors (with additional  $L_1$ norms on the
  $\beta$. $L_1$).
How will their results carry over to 0-1 loss?
\end{itemize}

\begin{figure}[tbhp] 
 \begin{adjustwidth}{-1.00in}{0in}
  \begin{subfigure}[b]{0.4\linewidth}
    \centering
   \includegraphics[totalheight=6cm]{./figs/phase-diagram-equivalence.png} 
\caption{{\bf The phase chage diagram.}}
\label{figure:phase-diagram-equivalence.png} 
    \vspace{4ex}
  \end{subfigure} 
  \begin{subfigure}[b]{0.6\linewidth}
    \centering
  \includegraphics[totalheight=6cm]{./figs/stepwise_lars_cv.png}
\caption{Stepwise lars. $p=200$, $n$ and $k$ variable. Model selected by cross-validation
}
\label{figure:stepwise_lars_cv.png}
  \end{subfigure} 
  \caption{}
\end{adjustwidth}
\end{figure}

\begin{figure}[tbhp]
\includegraphics[totalheight=6cm]{./figs/ranger.png}
\caption{ ranger}
\label{figure:ranger.png}
\end{figure}

>>>>

\newcommand{\cursedforest}{{\textsc{CursedForest}}
\newcommand{\mtry}{{\texttt mtry}}
\newcommand{\ntree}{{\texttt ntree}}

@@

The most popular random forest software packages are R-based, although current implementations do not scale well with
increasing feature size~\cite{Wright.and.Ziegle.2016}.  This is despite R now supporting long vectors (length greater
than $2^{31}-1$), and the \textsc{randomForest} implementation in R lending itself to parallelization to grow multiple
trees simultaneously~\cite{Liaw.and.Weiner.2002} and them combine them.  More suitable implementations for ``wide''
genomic data have been developed in other programming languages. \textsc{CloudForest} \cite{Bressler2015} written in Go,
achieves fast run times by effective use of the CPU cache, optimizing for different classes of features and efficiently
multi-threading.  Similarly, \textsc{Ranger}~\cite{Wright.and.Ziegle.2016} written in C++ with an R front-end is
specifically optimized for dealing with large data sets.

@@

Random forests apply the technique of bootstrap aggregation or bagging to decision trees.  The training data sets are
independently drawn, with replacement, from the joint distribution of $(X,Y)$ so that each set comprises $n$
$(p+1)$-tuples $(x_1,y_1),\ldots, (x_n,y_n)$. This is done $B$ times and a tree is grown on each of these $B$ samples.
The results are then combined in an appropriate way. In a classification problem we take the majority vote for the trees
over the $q=1,..,Q$ classes,
\begin{equation*}
{{h_f}}= \arg \max_q \left(\sum_b I(h(x;B_b)=q)\right).
\end{equation*}
In addition to growing the trees
on bootstrapped samples, each node in the tree is split on a randomly
selected subset of the variables.

@@

In the case of a regression tree, the average prediction error for an individual tree $h(X; B)$ is
\begin{equation}
PE_t = E_B E_{X,Y} (Y-h(X; B))^2.
\end{equation}
Assume that, for all $B$, the tree is unbiased, i.e., $EY= E_X h(X; B)$. Then
\begin{equation}
PE_f \leq \rho PE_t
\end{equation}
where $\rho$ is the weighted correlation between residuals $(Y-h(X;B))$ and $(Y-h(X;B^\prime))$ for independent $B,B^\prime$.  
So what is required for accurate random forest regression is (i) low correlation between residuals of differing tree in
the forest, and (ii) low prediction error for the individual trees \cite{Segal.2004}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% This section talks about the datasets
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Datasets}
\subsubsection{1000 Genomes Project}
We obtained the 1000 Genomes Project data as VCF files from their FTP site.  Each VCF file contains the variants for
every individual across one chromosome.  We use the phase 3 dataset which contains 2,504 individuals with approximately
81 million variant positions (i.e. a matrix of $2,504 \times 81$ million).  For compatibility with the majority of the
algorithms we demonstrate, we convert the original variant representations to double representations.  For example, we
store the absence of a variant (0\textbar0) as 0, a heterozygous
variant (1\textbar0 or 0\textbar1) as 1, and a
homozygous variant (1\textbar1) as 2.  This is not the only possible encoding.  \cite{Goldstein.et.al.2011} note that of
the encoding listed in table \ref{table:encodings}, the additive encoding will present the most difficulty for a random
forest model.

@@

\subsection{Parameter settings}
We consider the parameter setting for the random forest algorithm. We use the R notation from the \textsc{randomForest}
package \cite{Liaw.and.Weiner.2002} which incorporates the original Fortran code by Brieman and Cutler.  We incorporate
the advice of \cite{Liaw.and.Weiner.2002}, which we have found mirrors our own experience:
\begin{itemize}
\item \ntree\ the number of trees.  The number of trees necessary for good performance grows with the number of
  predictors.  \cite{Liaw.and.Weiner.2002} suggest that a high \ntree\ is necessary to get stable estimates of variable
  importance and proximity; however, even though the variable importance measures may vary from run to run, we note that
  it is possible for a random forest model to have a very poor fit and still have an accurate ranking of variable
  importance;
\item \mtry\ -- the number of variables considered at each split (if \mtry=$p$, we have a boosted decision
  tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger \mtry\ may give
  better performance;
\item the size and complexity of the individual trees is controlled in
  \textsc{randomForest} by setting \texttt{nodesize}, the minimum size of
  terminal nodes. It is controlled in Spark ML by setting
  \texttt{maxDepth}, the maximum depth of each tree in the
  forest\footnote{The Spark ML documentation \cite[]{Spark.2016} sets
    \texttt{maxDepth} to 4 in their classification example. This may
    give an estimate with a higher bias and, as such, may be a poor
    choice. See the discussion in \cite{Dietterich.2002} where it is
    shown that bagging may reduce bias as well as variance but gives a
    better result for low bias learners. \cite{Dietterich.2002} notes
    that if the boostrap replicate approximation were correct, then
    bagging would reduce variance without changing bias.} In the TCGA
  example the RandomForests code produced trees with depths of 30+.
\end{itemize}

@@

\marginpar{Does \cloudforest use Gini impurity or information gain? I
  thought it used infomation gain but Piotr mentions Gini}

@@

Firstly we look at \cursedforest horizontal scalability for a medium size dataset of 2.5 million variables and 5000
samples, by varying \mtry\ fraction and the number of CPU cores allocated to the execution. The raw results are
presented in table \ref{synthetictimingtable} and visualized in figure \ref{figure:synth} below. Regardless of the
number of cores used, \cursedforest\ displays approximately linear dependency between the execution time and the \mtry\
parameter (presented here as a fraction of the number of variables) [Fig 1 a).

@@

\begin{figure}[tbhp] 
 \begin{adjustwidth}{-1.00in}{0in}
    \centering
    \includegraphics[totalheight=6cm]{./figs/phase-diagram-equivalence.png} 
    \caption{{\bf The phase chage diagram.}}
    \label{figure:phase-diagram-equivalence.png} 
    \vspace{4ex}
  \end{adjustwidth}
\end{figure}

\section{not for the paper}
Notes:
\begin{itemize}
\item we try a small simulation, a regression problem with $\beta(1:k) \sim U(1,100)$ and $y= X\beta + \epsilon$, $\epsilon \sim  N(0,\sqrt{16})$
\item \cite{Donoho.and.Stodden.2006} evaluate variable selection by 
$$\frac{\Vert\hat{\beta}-\beta\Vert_2}{\Vert\beta\Vert_2}$$
See figure \ref{figure:stepwise_lars_cv.png} 
\item 
However there is no $\hat{\beta}$ in a tree model. We are using variable importance as
a measure of the ordering of $\hat{\beta}$ (i.e. we expect variable importance to be proportional to $\hat{\beta}$). Is this correct?
\item as RF is sort of greedy approximation to combinatorial approach, it may have some success in the dead zone
\item  \cite{Donoho.and.Stodden.2006} are considering problems with $L_2$ errors (with additional  $L_1$ norms on the
  $\beta$. $L_1$).
How will their results carry over to 0-1 loss?
\end{itemize}

\begin{figure}[tbhp] 
  \begin{adjustwidth}{-1.00in}{0in}
    \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[totalheight=6cm]{./figs/k.png}
      \caption{k}
      \label{figure:k.png}
      \vspace{4ex}
    \end{subfigure} 
    \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[totalheight=6cm]{./figs/n.png}
      \caption{n}
      \label{figure:n.png}
    \end{subfigure} 
    \caption{}
  \end{adjustwidth}
\end{figure}

\begin{figure}[tbhp] 
  \begin{adjustwidth}{-1.00in}{0in}
    \begin{subfigure}[t]{0.5\linewidth}
      \centering
      \includegraphics[totalheight=6cm]{./figs/stepwise_lars_cv.png}
      \caption{Stepwise lars. $p=200$, $n$ and $k$ variable. Model selected by cross-validation}
      \label{figure:stepwise_lars_cv.png}
      \vspace{4ex}
    \end{subfigure} 
    \begin{subfigure}[t]{0.5\linewidth}
      \centering
      \includegraphics[totalheight=6cm]{./figs/ranger.png}
      \caption{ranger, RBO}
      \label{figure:ranger.png}
    \end{subfigure} 
    \caption{}
  \end{adjustwidth}
\end{figure}
